---
output: github_document
---
# Introduction

This document looks into the relationship between population and area by country using the R programming language. The data used for this analysis is from this [wikipedia page.](https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population_density)

The ulimate goal of this analysis is to determine countries which are over/under populated when compared against countries with similar area.


```{r setup, include=FALSE}
#
# install.packages("rvest")
# install.packages("xml2")
# install.packages("magrittr")
library(rvest)
library(xml2)
library(magrittr)
library(tidyverse)
library(janitor)
library(lubridate)
library(broom)
library(knitr)
```

### Data preperation

Firstly, read in the data. This is farily easy to do in wikipedia. Following this [blog post](https://www.engineeringbigdata.com/web-scraping-wikipedia-world-population-rvest-r/) helped me.
```{r read_data}
html <- "https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population_density"
xpath <- '//*[@id="mw-content-text"]/div/table'
global <- html %>% read_html() %>%
  html_nodes(xpath = xpath) %>% html_table()


pop_raw <- global[[1]] %>% as_tibble() %>% clean_names() 
head(pop_raw)
```
Looking at the first 6 rows we can see that this data is in need of a good clean. In the next section of code I remove the commas and convert the column to their correct types.s
```{r clean_data, warning= FALSE}
pop <- pop_raw %>%
  mutate_at(c(
      "area_km2",
      "area_mi2",
      "population",
      "density_pop_km2",
      "density_pop_mi2"
    ),
   ~ str_replace_all(., ',', '') %>% as.numeric()
  ) %>%
  mutate(date = mdy(date),
         country_or_dependent_territory = str_remove(country_or_dependent_territory,'( *)\\[note .*')) %>%
  filter(!str_detect(country_or_dependent_territory,"World") ) %>%
  mutate_at(c("population","area_km2"),list(log = log))
head(pop)
```

This looks much neater.

```{r eval=FALSE, include=FALSE}
pop %>% ggplot(aes(log(population)))+geom_density()
```

### Data Modelling

Now lets look at the relationship. As the below graph shows, their seems to be a positive relationship between area and population. As area doubles, population doubles also. I've used log as it helps model the relationship better. Mathetically speaking I need to understand this better.

```{r explore_data, warning=FALSE}
pop %>% ggplot(aes(x = area_km2_log, y = population_log)) +
  geom_point() + geom_smooth(method = "lm")
```

Now let's build a model to help quanitfy this relationship.

```{r model_data}
pop_model <- lm(population_log ~ area_km2_log,pop)

summary(pop_model)
```

Looking at the summary, it appears there is a relationship between the two variables. The p value is less than 5% so we can reject the null hypothesis. The R squared value of `r round(summary(pop_model)$r.squared,2)` informs us that the relationship is strong but there is room for improving this model. 

Let's look at the residual plots to see if model follows assumptions.
```{r residual_plot}
par(mfrow=c(2,2))
plot(pop_model)
```

As this is only for fun, I'm happy with the residual plots and model statistics and feel it is suited for the purpose of this analysis.

Let's obtain residuals and convert them back into normal figures rather than log of population.

```{r}
pop_aug <- pop_model %>% augment(pop) %>%
  mutate(.resid = round(exp(.fitted)-population),
         .fitted = round(exp(.fitted)),
         error = round(population/.fitted,2))

```

### Conclusions

Looking at the below, Macau and Singapore are two of the countries who are over populated when compared against countries with similar area.

UK makes the top 20.

```{r inspect_model, eval=FALSE, include=FALSE}
pop_aug %>% ggplot(aes(x = population_log,y = .resid))+geom_point()
# As population increases we tend to over estimate the population in our model

pop_aug %>% ggplot(aes(y = .resid,x = .fitted))+geom_point()
pop_aug %>% ggplot(aes(x = .cooksd,y = .std.resid))+geom_point()
```
```{r, results = 'asis'}
view_pop <- pop_aug %>%  filter(population > 100000) %>% select(country_or_dependent_territory,area_km2,population,.fitted,.resid,error)

view_pop %>% arrange(desc(error)) %>% head(20) %>%  kable(caption = "Over Populated Countries",format = "markdown",format.args = list(big.mark = ",",scientific =F))
```

Similarly, we can see French Guiana and Western Sahara are two countries who are currently under populated when compared against countries of similar area.

Interesting to see Canada there.
```{r, results = 'asis'}
view_pop %>% arrange(error) %>% head(20) %>%  kable(caption = "Under Populated Countries",format = "markdown",format.args = list(big.mark = ",",scientific =F))
```

### Further Work

There is still work to do to improve model. The addingt the following may be good starting points.

* Average temperature
* GDP per capita
* Mountinous area
